Seasonality in governing dynamics at submesoscales in the Kuroshio Extension

### 2016-06-29 writeup/rocha_etal.tex
Some minor fixes by Sarah

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. While summertime
    re-stratification weakens submesoscale turbulence, it also enhances intertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the submesoscale surface kinetic energy and
    sea-surface height variance in late Summer/early Fall.
\end{abstract}

@@

Using the output of two $1/24^\circ$ and $1/48^\circ$ global
numerical simulations with embedded tides, we show that inertia-gravity waves undergo
a strong seasonality near the surface in the Kuroshio Extension region.
Interestingly, the seasonal cycle of inertia-gravity waves is out-of-phase
with the seasonal cycle of submesoscale turbulence. Consistent with previous studies,
deep late Winter mixed layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from a ECCO2 adjoint-method state estimate, which synthetized millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early Spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained,
its local maximum and RMS is smaller than $0.5f$.
Indeed, the seasonal cycle of vorticity and rate of strain are strong (figure~\ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late Winter/early Spring
than in late Summer/early Fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in Winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}a-c). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces 1.4 to 1.13 from hourly to daily-averaged.
 Thus wintertime submesoscale surface velocity is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
(figure \ref{fig3}g). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; figure \ref{fig3}h).

@@

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in Summer) and one with a very steep slope.
That the submesoscale dynamics in October is mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}j and \ref{fig3}k), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

@@

To better quantify the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$, which extends the 1/24$^\circ$
towards smaller scales (supplemental material).

@@

\section{Summary and Conclusion}
Our main finding is that in tidally forced global simulations the
near-surface upper-ocean inertia-gravity waves undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
submesoscale turbulence. \cite{dasaro1978} showed that velocity of linear internal waves
in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In Summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
change seasonally. In particular the baroclinic modes are
significantly more surface intensified in late Summer/early Fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore it is plausible to expect a strong seasonality in the near-surface
expression of internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in Winter, but those waves project on large
horizontal scales \citep[e.g, ][]{qi_etal1995}.

Our work adds to recent studies that presented modeling \cite{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, this study on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but is unlikely to be
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre; care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

@@

\subsection*{Spectra}
Kinetic energy and sea-surface height variance wavenumber spectra for the LLC2160
and LLC4320 simulations are shown in figure \ref{figS2_3}. As expected, and
consistent with the submesoscale regional study by \cite{capet_etal2008a}, the
higher-resolution simulation (LLC4320) essentially extends the spectra of the
lower-resolution simulation (LLC2160) towards smaller scale, demonstrating that
the higher resolution is resolving a wider array of scales. Between 20 and 100 km
there is linewidth agreement between the spectra of the two simulations.

>>>>

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides suggest a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-Winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late Winter/early Spring. While summertime
    re-stratification weakens submesoscale turbulence, it also enhances inertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the submesoscale surface kinetic energy and
    sea-surface height variance in late Summer/early Fall.
\end{abstract}

@@

Using the output of $1/24^\circ$ and $1/48^\circ$ global
numerical simulations with embedded tides, we show that inertia-gravity waves undergo
a strong seasonality near the surface in the Kuroshio Extension region.
Interestingly, the seasonal cycle of inertia-gravity waves is out-of-phase
with the seasonal cycle of submesoscale turbulence. Consistent with previous studies,
deep late Winter mixed layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.

@@

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from an ECCO2 adjoint-method state estimate, constrained to millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 5-month long output used by
\citet{rocha_etal2016}.

@@

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early Spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
Fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained;
its local maximum and RMS are both smaller than $0.5f$.
Indeed, the seasonal cycle of vorticity and rate of strain are strong (figure~\ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late Winter/early Spring
than in late Summer/early Fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in Winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ significantly increases the wintertime RMS vorticity
and strain by about 40$\%$.

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (see figures \ref{fig3}a-c). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, and residual velocities, although
 the vorticity skewness reduces from 1.4 to 1.13 from hourly to daily-averaged.
 The April results are characteristic of Winter, indicating that wintertime submesoscale
 surface velocity is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
(figure \ref{fig3}g). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; figure \ref{fig3}h).

@@

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in Summer) and one with a very steep slope.
That the submesoscale dynamics in October are mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly and
residual fields
(figure \ref{fig3}j and \ref{fig3}k), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

@@

To better quantify the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.  Before calculating the spectra, time-mean and spatial linear
trends were removed, and the resulting fields were multiplied by a two-dimensional Hanning ``spectral window''.
The two-dimensional spectra were averaged azimuthly \citep[e.g., ][]{rocha_etal2016} .
We discuss only spectra for the 1/48$^\circ$ simulation, which extends the 1/24$^\circ$ simulation
towards smaller scales (supplemental material).

@@

\section{Summary and Conclusion}
Our main finding is that in global simulations with embedded tides the
near-surface upper-ocean inertia-gravity waves in the Kuroshio Extension undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
submesoscale turbulence. \cite{dasaro1978} showed that velocity of linear internal waves
in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In Summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
changes seasonally. In particular the baroclinic modes are
significantly more surface intensified in late Summer/early Fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore it is plausible to expect a strong seasonality in the near-surface
expression of internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in Winter, but those waves project on large
horizontal scales \citep[e.g, ][]{qi_etal1995}.

Our work adds to recent studies that presented modeling \citep{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean. Of course, this study focuses on a single patch of ocean in the
vicinity of the Kuroshio Extension, which may be typical of mesoscale-rich
subtropical regions, but is unlikely to be
representative of other regions such as low eddy kinetic energy eastern boundary currents
and the middle of the subtropical gyre; care must be taken to avoid overgeneralizing
these results.  We plan to report on the geographic variability of submesoscale
 seasonality in a future study.

@@

\subsection*{Spectra}
Kinetic energy and sea-surface height variance wavenumber spectra for the LLC2160
and LLC4320 simulations are shown in figure \ref{figS2_3}. As expected, and
consistent with the submesoscale regional study by \cite{capet_etal2008a}, the
higher-resolution simulation (LLC4320) essentially extends the spectra of the
lower-resolution simulation (LLC2160) towards smaller scale, demonstrating that
the higher resolution is resolving a wider array of scales. Between 20 and 100 km
the spectra of the two simulations are visually indistinguishable.
